Making ice sheet models \\ scale properly

### 2023-01-06 talk/slides.tex
many edits

\begin{itemize}
\item before numerical modeling \dots \emph{what} are we modeling?
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
\end{itemize}

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\vspace{13mm}}
\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
\only<5>{\par {\scriptsize but an ice sheet is \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{basic facts}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon!
    \end{itemize}
\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of physics!
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}
\item now, a simple definition of ``ice sheet model'' \dots
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

    \item[$\circ$] \emph{bed elevation} $b(x,y)$
    \end{itemize}
\item two outputs:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x,y)$
    \item[$\circ$] \emph{ice velocity} $\bu(t,x,y,z)$
    \end{itemize}
\item {\small $\begin{pmatrix} \text{climate} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{frame}{notation}

\begin{itemize}
%\item but this is a mathematics talk \dots notation! equations!
\item the data $a(t,x,y)$, $b(x,y)$ is defined on fixed time and space domains:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$

@@

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] $\Gamma_s(t),\Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{conservation}

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary conservation} problem
    \begin{itemize}
    \item[$\circ$] mass conservation
    \item[$\circ$] momentum conservation
    \item[$\circ$] energy conservation (\emph{ignored here})
    \end{itemize}
\item key idea: conservation of mass means both conserve mass in interior (incompressible) and at surface (SKE)
\end{itemize}
\end{frame}

\begin{frame}{strong form of ice sheet model}

\begin{itemize}
\item nonlinear complementarity problem coupled to Stokes:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\bu\cdot \bn = 0 \quad \& \quad \btau_b - \bbf(\bu) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}
\end{align*}

    \begin{itemize}
    \item viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

@@

\section{Stokes solver scaling}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item current-technology ice sheet models mostly use \alert{explicit time stepping}, \alert{non-optimal stress-balance solvers}, and shallow assumptions
\item \alert{coming soon} from current research:
   \begin{itemize}
   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[$\circ$] ML emulators (Jouvet et al.~2021)
   \end{itemize}
\item scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\footnotesize
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021). \emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
\item A.~L{\"o}fgren, J. Ahlkrona, \& C. Helanow (2022). \emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
\end{itemize}
\end{frame}

>>>>

\begin{itemize}
\item before numerical modeling \dots \emph{what} are we modeling?
\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
\end{itemize}

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\vspace{13mm}}
\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
\only<5>{\par {\scriptsize \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{basic facts about glaciers}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item the geometry and velocity evolve \emph{in contact with its climate},
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating (\emph{ice shelf})
    \item[$\circ$] calving into the ocean
    \end{itemize}
\item and the \emph{topography} on which it sits
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{simplifications}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming model performance analysis, I will ignore quite a bit of physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})} for a model which does \emph{not} make these simplifications}
\item<2> {\footnotesize next, a simple definition of ``ice sheet model''}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

    \item[$\circ$] \emph{bed elevation} $b(x,y)$
    \end{itemize}
\item two outputs:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x,y)$
    \item[$\circ$] \emph{ice velocity} $\bu(t,x,y,z)$
    \end{itemize}

\bigskip
\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{frame}{notation for the simplified model}

\begin{itemize}
\item the \alert{data} $a(t,x,y)$, $b(x,y)$ \alert{is defined on a fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$

@@

\vspace{-2mm}
\end{itemize}
\end{frame}

\begin{frame}{mass and momentum conservation}

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored in this talk}
    \end{itemize}
\item a \emph{key idea} is that conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}
    \end{itemize}
\item where $\Gamma_s(t)$ and $\Gamma_b(t)$, subsets of $\partial \Lambda(t)$, denote the surface and base of the ice
\end{itemize}
\end{frame}

\begin{frame}{free boundary problem}

\begin{itemize}
\item but ice sheet evolution is a \alert{free-boundary} problem for conserved quantities in a (thin) fluid layer
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the fixed map-plane domain $\Omega\subset \RR^2$, glacier \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

\bigskip
\item<2> {\footnotesize for more on this free-boundary perspective, see Jouvet \& Bueler (2012) and Bueler (2021)}
\end{itemize}
\end{frame}

\begin{frame}{simplified ice sheet model (strong form)}

\begin{itemize}
\item nonlinear complementarity problem \onslide<2->{coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
\end{align*}

    \begin{itemize}
    \item $\bu|_s=0$ where no ice
    \item<2-> viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{ice thickness and ``mass continuity equation'' view}

FIXME
\end{frame}

@@

\section{on Stokes solver scaling}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
\item<3-> \alert{coming soon} from current research:
   \begin{itemize}
   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[$\circ$] ML emulators (Jouvet et al.~2021)
   \end{itemize}
\item<4-> scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

@@

{\scriptsize
\begin{itemize}
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet \& E. Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math. 72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021). \emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
\item A.~L{\"o}fgren, J. Ahlkrona, \& C. Helanow (2022). \emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
\end{itemize}
}
\end{frame}
